---
title       : The caret package
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
output: html_document
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
require(knitr)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig_010_caret/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## The caret R package

<img class=center src=https://raw.githubusercontent.com/DataScienceSpecialization/courses/master/assets/img/08_PredictionAndMachineLearning/caret.png height=400>

[http://caret.r-forge.r-project.org/](http://caret.r-forge.r-project.org/)

---

## Caret functionality

* Some preprocessing (cleaning)
  * preProcess - clean data and setup the features.
* Data splitting
  * createDataPartition
  * createResample
  * createTimeSlices
* Training/testing functions
  * train
  * predict
* Model comparison
  * confusionMatrix - info on how well the model works on new data sets.

---

## Machine learning algorithms in R

* Linear discriminant analysis
* Regression
* Naive Bayes
* Support vector machines
* Classification and regression trees
* Random forests
* Boosting
* etc. 

These are written by different developers and so they have different interfaces.

---

## Why caret? 

Consider this list of different prediction algorithms from different packages. For each algorithm we have created an object called `obj`.  Notice that each object has a different class, and that each prediction requires different parameters to be passed to it.

<img class=center src=https://raw.githubusercontent.com/DataScienceSpecialization/courses/master/assets/img/08_PredictionAndMachineLearning/predicttable.png height=250>

The caret package provides a unifying framework which allows you to predict using just one function and without having to specify all the options.

http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf


--- 

## SPAM Example: Data splitting

```{r loadPackage}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
```

- `p=0.75`: The training set will use 75% of the data.
- `testing <- spam[-inTrain,]`: The testing subset is just the data that isn't in `inTrain`.

--- 

## SPAM Example: Fit a model
Here we fit a model using the caret package.
 - We use the `train()` function.
 - `~.` indicates that we want to use all the other variables to predict.

```{r training, dependson="loadPackage",cache=TRUE}
set.seed(32343)
modelFit <- train(type ~.,data=training, method="glm")
modelFit
```

The `train()` function will test the model (in this case, a glm) in a variety of ways.
- In this case, it decided to use `Resampling: Bootstrapped` with 25 replicants.
    - It also corrects for the potential bias that might come from bootstrapping.

--- 

## SPAM Example: Final model

Here we look at the fitted values for our model by using the command `<name of model>$finalModel`.

```{r finalModel, dependson="training",cache=TRUE}
modelFit <- train(type ~.,data=training, method="glm")
modelFit$finalModel
```

--- 

## SPAM Example: Prediction
Then we can predict our test data like so:

```{r predictions, dependson="training",cache=TRUE}
predictions <- predict(modelFit,newdata=testing)
predictions
```

--- 

## SPAM Example: Confusion Matrix
The confusion matrix will help us evaluate how well our model fit works.  
We need to pass it two parameters:

    - The predictions from your model fit.
    - The outcome (output variable) from the testing sample.
```{r confusion, dependson="predictions",cache=TRUE}
confusionMatrix(predictions,testing$type)
```

---

## Further information

* Caret tutorials:
  * [http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf](http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf)
  * [http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf](http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf)
* A paper introducing the caret package
  * [http://www.jstatsoft.org/v28/i05/paper](http://www.jstatsoft.org/v28/i05/paper)