---
title       : Data slicing
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
require(knitr)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig_011_data_slicing/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## SPAM Example: Data splitting

Here we use the `createDataPartition()` function to create a training set and a testing set.
- We split the data by the `spam$type` column, 
    - which is a factor of two [spam; nonspam].
- We choose `p=0.75` to specify that 75% of the data is allocated to the training set.
- The `inTrain` variable is now assigned an indicator function which can be used to subset our two sets.

```{r loadPackage}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
```

---

## SPAM Example: K-fold
For a cross-validation set, we use `createFolds()`.
- We're splitting on `spam$type`.
- We'll have 10 folds.
- `list=TRUE` means that it will return each set of indices corresponding to a particular fold as a list.
- `returnTrain` tells it to return the training set or the test set.

```{r kfold,dependson="loadPackage"}
set.seed(32323)
folds <- createFolds(y=spam$type,k=10,
                             list=TRUE,returnTrain=TRUE)
sapply(folds,length)  ## Check the lengths of our folds.
folds[[1]][1:10] ## Check the indices (row numbers) of the samples.
```
That last command shows us that the folds were split while preserving  the original order of the data set; they weren't re-ordered. That is, the first 4141 samples are in the first fold, and the very first sample is still the first on the list.

This means that we can easily know which of the indices are corresponding to the training and test sets within those k folds. 

---

## SPAM Example: Return test
Here we return the test set. 

```{r kfoldtest,dependson="loadPackage"}
set.seed(32323)
folds <- createFolds(y=spam$type,k=10,
                             list=TRUE,returnTrain=FALSE)
sapply(folds,length)
folds[[1]][1:10]
```

You can see that the folds are smaller because the test set is only 25% of the data.

---

## SPAM Example: Resampling
`createResample()` is used for resampling or bootstrapping instead of k-fold cross validation. 
 - Instead of a list we could have a matrix.
- Resampling uses replacement, which means we could get duplicated samples.

```{r resample,dependson="loadPackage"}
set.seed(32323)
folds <- createResample(y=spam$type,times=10,
                             list=TRUE)
sapply(folds,length)
folds[[1]][1:10]
```

Notice that in fold #1, sample #3 was used three times, and #5 twice.

---

## SPAM Example: Time Slices
With time slices we can specify a number of consecutive samples to take as the training set, and a number of subsequent consecutive samples to skip as the test set. That way we can use the time-varying component of the set to predict.

- `initialWindow=20` : 20 samples per window for the training set.
- `horizon=10` : The next 10 samples for the test set.

```{r time,dependson="loadPackage"}
set.seed(32323)
tme <- 1:1000
folds <- createTimeSlices(y=tme,initialWindow=20,
                          horizon=10)
names(folds)
folds$train[[1]]
folds$test[[1]]
```

---

## Further information

* Caret tutorials:
  * [http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf](http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf)
  * [http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf](http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf)
* A paper introducing the caret package
  * [http://www.jstatsoft.org/v28/i05/paper](http://www.jstatsoft.org/v28/i05/paper)